{fairmetrics}: An R package for fairness evaluation metrics with confidence intervals

### 2025-05-28 paper.Rmd
adding jesse's comments from the md file to the rmd file.
Using HTML comments for Rmd

{fairmetrics} is an R package designed to evaluate the fairness of machine learning models through a range of specialized metrics for which a model can be classified as "fair". It supports fairness assessments of popular group-based criteron, such as independence, separation, sufficency and others. The package enables statistical inference on fairness metrics through calculation of bootstrap confidence intervals (CIs). In addition, {fairmetrics} offers convenient wrapper functions to compute multiple metrics simultaneously and includes datasets derived from the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical] for illustrating its use.

@@

@@

An ML model can be evaluated as being "fair" through three major criteron: group fairness, individual fairness and causal fairness. Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions  [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. {fairmetrics} focuses on calculating group fairness metrics as they are commonly used in biomedical settings. The groups in question are most often defined by protected attributes, such as age or race [@mehrabiSurveyBiasFairness2021].

Group fairness criteria are commonly categorized into three main types: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. Independence requires that an ML model's predictions be statistically independent of the protected attribute. Separation demands that the model's predictions be independent of the protected attribute conditional on the true outcome class (i.e., within the positive and negative classes). Sufficiency requires that, given a model's prediction, the likelihood of the true outcome is independent of the protected attribute—aiming to equalize error rates across groups for similar prediction score. The {fairmetrics} package computes a range of group fairness metrics along with bootstrap-based confidence intervals. These metrics are grouped below according to the three core fairness frameworks described above.

@@

@@

-   **Predictive Equality:** Compares false positive rates (FPR) between groups, ensuring that no group is disproportionately flagged as positive when the true outcome is negative.

@@

The {fairmetrics} package requires that a model has been trained and validated with data that has been split appropriately. The primary required input is a data frame or tibble which containing the model predictions, true outcomes, and protected attributes. \hyperref[workflow]{Figure ~\ref*{workflow}}
 shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.

@@

More concretely, an example of how to use the {fairmetrics} package is shown below. The example uses the `mimic_preprocessed` dataset, which is a preprocessed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].

Consider a random forest model trained to predict the risk of heart failure in patients. Suppose we are interested in evaluating the model's fairness with respect to patient gender with predictive equality.

@@

Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels}[@wisniewski2022fairmodels]. The differences between {fairmetrics} and these other packages is twofold. The primary difference between is that {fairmetrics} allows for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated. Additionally, the {fairness} package has fewer dependencies and a lower memory footprint, making the for a more environment agnostic tool that can be used with modest hardware. \hyperref[tab:memory_usage]{Table~\ref*{tab:memory_usage}} shows the memory comparison when loading each library.

\begin{table}[ht]
\centering
\begin{tabular}{l r}
\hline
\textbf{Package} & \textbf{Memory (MB)} \\
\hline
fairmodels  & 17.02  \\
fairness    & 117.61 \\
fairmetrics & 0.05   \\
\hline
\end{tabular}
\caption{Memory usage of {fairmetrics} vs similar packages (MB)}
\label{tab:memory_usage}
\end{table}

>>>>

<!--- 
Rewritten by Jesse
--->
Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example an dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

<!---
ianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.  We can of course use it as an example but this isn't the need for the package. Can you rewrite this to just be about the general use of fairness?  i.e., point 1 - ml increasing integrated into different areas, point 2 - give a few examples, point 3 - need for the package in light of other work.  I don't think interpetable is the correct word here, the key difference is allowing for uncertainty quantification, we also provide ratio + difference estimates, i dont think R fairness package does, I would emphasize   
--->

@@

<!---
# REMOVED: not relevant to the package
An ML model can be evaluated as being "fair" through three major criterion: group fairness, individual fairness and causal fairness. Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions  [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. {fairmetrics} focuses on calculating group fairness metrics as they are commonly used in biomedical settings. The groups in question are most often defined by protected attributes, such as age or race [@mehrabiSurveyBiasFairness2021].
--->

Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. The {fairmetrics} package computes a range of group fairness metrics together with bootstrap-based confidence intervals for uncertainty quantification.  The metrics implemented in the package are briefly described below.

<!---
1. Jianhui - The scope of the package is not defined in the above paragprah so one would not understand the definitons below (i.e., "positive classification" is never defined).  Specifically, you need to say you consider binary classification and a binary protected attribute - I do not believe your package handles more than.  Explain why this is done and then update the definitions below to reflect this.

2. Jianhui - why do you use bootstrap for this package rather than IF?  Is it because you don't have IF for all metrics?

3. Jianhui - please update the initial sentences describing the 3 categories to be more intuitive, akin to what we have in the paper.  
--->

@@

<!---
Use consistent language throughout - alternating between checks/compares/focuses/assess/measure/etc is confusing as all functions do the same thing.  Suggest use "Compares" for all. 
--->

@@

<!---
## Do you need the acronyms for FNR and FPR? If you don't use them later, they don't need to be defined.
--->
-   **Predictive Equality:** Compares false positive rates (FPR) between groups, ensuring that no group is disproportionately flagged as positive when the true outcome is negative.

@@

The input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attributes (attribute? doesn't package only handle 1?). \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.

@@

<!---
Edited by Jesse
--->
A simple example of how to use the {fairmetrics} package is shown below. The example uses the `mimic_preprocessed` dataset, which is a pre-processed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].

<!---
1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.

2. Ben - suggest you show more of the functions in the package - say that, while the choice of metric is context dependent, we show all metrics to show the full range of the package. Also include interpretation of the various results.  Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.  
---->

@@

<!---
## Both - you are missing this one: https://mlr3fairness.mlr-org.com
## Do these metrics give difference and ratio based criteria? Are there any other differences? 
--->
Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels} [@wisniewski2022fairmodels]. The differences between {fairmetrics} and these other packages is twofold. The primary difference between is that {fairmetrics} allows for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated. Additionally, in contrast to the {fairmodels} and {fairness} packages, the {fairmetrics} package has zero library dependencies and a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems. \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.

\begin{table}[ht]
\centering
\begin{tabular}{l r r}
\hline
\textbf{Package} & \textbf{Memory (MB)} & \textbf{Dependencies} \\
\hline
fairmodels  & 17.02  & 29\\
fairness    & 117.61 & 141\\
fairmetrics & 0.05   & 0\\
\hline
\end{tabular}
\caption{Memory usage and dependencies of {fairmetrics} vs similar packages (MB)}
\label{tab:memory_dep_usage}
\end{table}
